---
title: "Clinical Metadata for curatedMetagenomic Projects"
output: html_notebook
---
## BACKGROUND 
The https://waldronlab.io/curatedMetagenomicData/ is a curated repository of organized metagenomic files from a group of cohorts with associated clinical data.

We decided to use this data to look into CRC patients, their metagenomic data, and run Taxa- and BGC- on such files to expand the scope of our research into CRC. 

### Is there a connection between CRC and a particular BGC?
Connection between certain taxa and CRC is presently tenuous, and various depending on the study taken into consideration

### How much clinical data is present in CRC studies?

The "curatedMetagenomicData.tsv" is a file in https://waldronlab.io/curatedMetagenomicData/ that supposedly contains patient/sample level metagenomic values. 

Here I take a look at "how much metadata" is actually there
```{r}
# dependencies
library(tidyverse)
library(readxl)

# upload files and find levels of study conditions
curatedmd <- read_tsv("raw-data/curatedMetagenomicData.tsv",show_col_types = FALSE)
curatedmd$study_condition %>% unique()
```
CRC --> Colon Rectal Cancer, it is the only "lable" used for such disease

What are the studies with associated CRC lable?
```{r}
curatedmd_CRCnames <- curatedmd %>%
  filter(study_condition == "CRC") %>%
  group_by(study_name) %>%
  summarise(count = n()) %>% 
  pivot_wider(names_from = study_name,values_from = count) %>% names()

curatedmd_CRCnames
```

Only 11 studies have been tagged as CRC studies.
How much *clinical* metadata is present in these studies?
```{r}
var <- names(curatedmd)[-c(1:3)]

toStringUnique <- function(vector) {
  toString(unique(vector))
}

varSummaryTbl <- curatedmd %>%
  filter(study_name %in% curatedmd_CRCnames) %>%
  group_by(study_name) %>%
  summarise(across(.col = var, .fns = toStringUnique)) %>%
  ungroup() %>% 
  select(!c(12:22)) %>% #remove columsn that are not clinical annotation values
  
  pivot_longer(cols = !c(study_name),names_to = "Variables") %>%
  pivot_wider(names_from = study_name)

varSummaryTbl
```
By looking a the table above, there is a very little amount of clinical data here. There is no reference to responders and non-responders, for example, to medications, and many of the other clinical parameters are missing.

### Presence of internal control cohorts in CRC studies
clinical observational studies with internal control cohorts are considered better for biological interpretation. Therefore, I checked if any of the CRC studies have associated ctr samples, to include in the downstream analysis.

Below is a complete list of studies that include internal control cohorts
```{r message=FALSE, warning=FALSE}

df1 <- curatedmd %>% 
  group_by(study_name,study_condition) %>%
  summarise(count = n()) %>%
  select(study_name)
vec <- df1$study_name
studyNamesWithCtr <- vec[ave(seq_along(vec), vec, FUN = length)>1] %>% unique()

studyNamesWithCtr
```
sanity check to make sure the list is accurate
```{r message=FALSE, warning=FALSE}
df %>%
  group_by(study_name,study_condition) %>%
  summarise(count = n()) %>%
  filter(study_name == studyNamesWithCtr[1])
```

Lastly, check if any of these are CRC studies
```{r message=FALSE, warning=FALSE}
studiesCRCwithCtr <- df %>%
  group_by(study_name,study_condition) %>%
  summarise(count = n()) %>%
  filter(study_name %in% studyNamesWithCtr[1:length(studyNamesWithCtr)]) %>%
  filter(study_condition == "CRC")

write_csv(studiesCRCwithCtr,"curated-data/studiesCRCwithCtr.csv")

studiesCRCwithCtr
```
🔴 **CONCLUSIONS**: All 11 of the CRC studies have internal cohorts as control

## Explore metadata directly from the publication
The lack on clinically-relevant metadata in the curatedMetagenomicData doesn't necessary mean that this data is missing. **I looked into each of the original publications** to see if they contain any annotation we can incorporate in the our database.

Below is a manual annotation ofregarding the presence or absence of metadata in each study

**FengQ_2015**              --> Original Article, patient-level data found 	

**GuptaA_2019**             --> Original Article, patient-level data found

**HanniganGD_2017**         --> Original Article, patient-level data found

**ThomasAM_2018ab_2019c**	  --> meta-analysis of five publicly available datasets (to dig in)
  - new ita cohort 1 (metadata should be in curatedMetagenomicData) --> not seen under "ThomasAM_2018a"
  - new ita cohort 2 (metadata should be in curatedMetagenomicData) --> not seen under "ThomasAM_2018a"
  - Validation Cohort1 is available in the European Nucleotide Archive under the study identifier PRJEB27928
  - Validation Cohort2 is available in the DNA data bank of Japan databases under the accession No. DRA006684

**VogtmannE_2016**    		--> Original Work, patient-level data not available	

**WirbelJ_2018**			    --> Meta-analysis:
  - four published studies which used fecal shotgun metagenomics to characterize CRC patients compared to healthy controls
  - An additional fifth population was generated from samples collected in Germany (herein abbreviated as DE); a subset of samples from this patient collective were published previously
  - Metadata for all the studies in this meta-analyysis are found in the original paper and downloaded here

**YachidaS_2019**		       --> Found, with additional metabolomic

**YuJ_2015**	             --> data are available only aggregated

**Conclusions**: I was able to pull clinical data from 5 out of the 8 studies. I still need to work on ThomasAM data that seem to be scatter in different location.

## WirbelJ_2018
A folder in raw-data ewas created containn relaveant published tables that need to be curated for data exploration
```{r}
# Table from Supplementary
WirbelJ_2018_SupplDataMOESM3sh3_1 <- read_excel("raw-data/WirbelJ_2018_PMID30936547/41591_2019_406_MOESM3_ESM.xlsx",sheet = 3, skip = 18)

# Adjust data type
WirbelJ_2018_SupplDataMOESM3sh3_2 <- WirbelJ_2018_SupplDataMOESM3sh3_1 %>%
  mutate(across(names(WirbelJ_2018_SupplDataMOESM3sh3_1[,c(4,14)]), as.numeric)) %>%
  mutate(across(names(WirbelJ_2018_SupplDataMOESM3sh3_1[,c(3,5:13,15,16)]), as.factor))

# I have now a well formatted table
WirbelJ_2018_SupplDataMOESM3sh3_2
write_csv(WirbelJ_2018_SupplDataMOESM3sh3_2,"curated-data/WirbelJ_2018_SupplDataMOESM3sh3_2.csv")
```

## GuptaA_2019
A folder in raw-data ewas created containn relaveant published tables that need to be curated for data exploration
```{r}
# I confirmed there is lots of data in the published dataset
# curate the table ready for merge

## Table from Supplementary Table 1
GuptaA_2019_SupplData1_1 <- read_excel("raw-data/GuptaA_2019_PMID31719139/mSystems.00438-19-st001.xlsx", skip = 1)
n<-dim(GuptaA_2019_SupplData1_1)[1]
GuptaA_2019_SupplData1_2 <- GuptaA_2019_SupplData1_1[1:(n-2),] #remove last two empty rows

# Adjust data type
GuptaA_2019_SupplData1_3 <- GuptaA_2019_SupplData1_2 %>%
  mutate(across(names(GuptaA_2019_SupplData1_2[,c(2,4,8:13)]), as.factor))

# I have now a well formatted table
GuptaA_2019_SupplData1_3
write_csv(GuptaA_2019_SupplData1_3,"curated-data/GuptaA_2019_SupplData1_3.csv")
```

## HanniganGD_2017
A folder in raw-data ewas created containn relaveant published tables that need to be curated for data exploration
```{r}
# patient metadata was obtained from github Hannigan_CRCVirome_mBio_2018/data/metadata/303030-metadata.tsv

## Table from Supplementary
HanniganGD_2017_SupplData1_1 <- read_csv("raw-data/HanniganGD_2017_PMID30459201/meta.csv",show_col_types = FALSE)

# Adjust data type
HanniganGD_2017_SupplData1_2 <- HanniganGD_2017_SupplData1_1 %>%
  mutate(across(names(HanniganGD_2017_SupplData1_1[,c(1,5,6,8,9)]), as.factor))

# I have now a well formatted table
HanniganGD_2017_SupplData1_2
write_csv(HanniganGD_2017_SupplData1_2,"curated-data/HanniganGD_2017_SupplData1_2.csv")

```
## YachidaS_2019
A folder in raw-data was created contain relaveant published tables that need to be curated for data exploration. 

This publication provides both clinical data as well as metabolomic data (count). Two tables, each containing cliincal data, were merged into one

```{r}
## First Table comes from Sheet 4 (Clinical data)
YachidaS_2019_SupplData2_1 <- read_excel("raw-data/YachidaS_2019_PMID31171880/41591_2019_458_MOESM3_ESM.xlsx",sheet = 4,skip = 2)

newHeader <- str_replace(string = names(YachidaS_2019_SupplData2_1),pattern = "[[:space:]]",replacement = "")
names(YachidaS_2019_SupplData2_1) <- newHeader

YachidaS_2019_SupplData2_2 <- YachidaS_2019_SupplData2_1 %>%
  mutate(across(names(YachidaS_2019_SupplData2_1[,2:3]), as.factor)) %>%
  mutate(across(names(YachidaS_2019_SupplData2_1[,5]), as.factor)) %>%
  mutate(across(names(YachidaS_2019_SupplData2_1[,6:8]), as.numeric)) %>%
  mutate(across(names(YachidaS_2019_SupplData2_1[,9]), as.factor))

# I have now a well formatted table
YachidaS_2019_SupplData2_2

## Second Table comes from Sheet 5 (metabolomics)
YachidaS_2019_SupplData3_1 <- read_excel("raw-data/YachidaS_2019_PMID31171880/41591_2019_458_MOESM3_ESM.xlsx",sheet = 5, skip = 2)

newHeader <- str_replace(string = names(YachidaS_2019_SupplData3_1),pattern = "[[:space:]]",replacement = "")
names(YachidaS_2019_SupplData3_1) <- newHeader

YachidaS_2019_SupplData3_2 <- YachidaS_2019_SupplData3_1 %>%
  mutate(across(names(YachidaS_2019_SupplData2_1[,1]), as.character)) %>%
  mutate(across(names(YachidaS_2019_SupplData3_1[,2:3]), as.factor)) %>%
  mutate(across(names(YachidaS_2019_SupplData3_1[,5]), as.factor)) %>%
  mutate(across(names(YachidaS_2019_SupplData3_1[,6:8]), as.numeric)) %>%
  mutate(across(names(YachidaS_2019_SupplData3_1[,9]), as.factor))

# I have now a well formatted table
YachidaS_2019_SupplData3_2

# bind
YachidaS_2019_SupplData1_1 <- bind_rows(YachidaS_2019_SupplData2_2,YachidaS_2019_SupplData3_2)
YachidaS_2019_SupplData1_1
write_csv(YachidaS_2019_SupplData1_1,"curated-data/YachidaS_2019_SupplData1_1.csv")
```

This study also comes with **metabolomic** data, per patient!

```{r}
## This is a Metabolomic Table
YachidaS_2019_SupplData13_1 <- read_excel("raw-data/YachidaS_2019_PMID31171880/41591_2019_458_MOESM3_ESM.xlsx",sheet = 18,skip = 3)
names(YachidaS_2019_SupplData13_1)[1] <- "metabolite" 

YachidaS_2019_SupplData13_2 <- YachidaS_2019_SupplData13_1 %>%
  pivot_longer(cols = !c("metabolite"), names_to = "Subject_ID") %>%
  select(Subject_ID, everything())

# I have now a well formatted table
YachidaS_2019_SupplData13_2
write_csv(YachidaS_2019_SupplData13_2,"curated-data/YachidaS_2019_SupplData13_2.csv")
```

Merge metabolomic table with clinical annotation using Subject_ID as reference 

```{r}
YachidaS_2019_SupplData1_2 <- inner_join(YachidaS_2019_SupplData1_1,YachidaS_2019_SupplData13_2,by="Subject_ID")
YachidaS_2019_SupplData1_2

write_csv(YachidaS_2019_SupplData1_2,"curated-data/YachidaS_2019_SupplData1_2.csv")


```

Visualization of relevant metabolites (higher mean value than 10)

```{r}
# Distribution of matabolites
relevantMetab <- YachidaS_2019_SupplData1_2 %>%
  select(metabolite,value) %>%
  group_by(metabolite) %>%
  summarise(mean = mean(value)) %>%
  filter(mean >= 10) %>% 
  pivot_wider(names_from = metabolite, values_from = mean) %>% names()
  
p1 <- YachidaS_2019_SupplData1_2 %>%
  filter(metabolite %in% relevantMetab) %>%
  mutate(valueLOG10 = log10(value)) %>%
  
  ggplot(aes(x=metabolite,y=valueLOG10)) +
  geom_boxplot(outlier.shape = NA) +
  theme_bw()+
  theme(axis.text.x = element_blank())

p1
```
Do any of these metabolites correlate with tumor Stage?

```{r message=FALSE, warning=FALSE}
# Does any of these metabolites correlate with tumour stage?
p2 <- YachidaS_2019_SupplData1_2 %>%
  filter(metabolite %in% relevantMetab) %>%
  mutate(valueLOG10 = log10(value)) %>%
  filter(metabolite == "C06423_Octanoate") %>%

  ggplot(aes(x=Stage,y=valueLOG10)) +
  geom_violin() +
  geom_jitter(size = 1)

p2

# Does any of these metabolites correlate with Age?
p3 <- YachidaS_2019_SupplData1_2 %>%
  filter(metabolite %in% relevantMetab) %>%
  mutate(valueLOG10 = log10(value)) %>%
  filter(metabolite == "C06423_Octanoate") %>%

  ggplot(aes(x=Age,y=valueLOG10)) +
  geom_point() +
  geom_smooth(method='lm',na.rm=TRUE)

p3
```

## FengQ_2015 
A folder in raw-data ewas created containn relaveant published tables that need to be curated for data exploration

```{r message=FALSE, warning=FALSE}

# Table from Supplementary material of the manuscripot
mdFengQ_2015_SupplData1_1 <- read_excel("raw-data/FengQ_2015_PMID25758642/41467_2015_BFncomms7528_MOESM1193_ESM.xlsx",
                                        skip = 2)
                                        
newHeader <- str_replace(string = names(mdFengQ_2015_SupplData1_1),pattern = "\\.\\.\\..+",replacement = "")

for (i in 1:length(newHeader)){
  hold <- newHeader[i]
  if (newHeader[i+1] == ""){
    newHeader[i+1] <- hold
  }
  # break check
  if (i==51) break
}

names(mdFengQ_2015_SupplData1_1) <- paste0(newHeader,"_",mdFengQ_2015_SupplData1_1[1,])
mdFengQ_2015_SupplData1_2 <- mdFengQ_2015_SupplData1_1[-1,]

names(mdFengQ_2015_SupplData1_2)[48] <- "MLG classifier_Type_Probability of carcinoma"
names(mdFengQ_2015_SupplData1_2)[50] <- "MLG classifier_Type_Probability of advanced adenoma"
names(mdFengQ_2015_SupplData1_2)[52] <- "MLG classifier_Type_Probability of advanced adenoma add age"
names(mdFengQ_2015_SupplData1_2)[1] <- "Sample ID"

names(mdFengQ_2015_SupplData1_2) <- str_replace_all(string = names(mdFengQ_2015_SupplData1_2),pattern = "[[:space:]]|\\(|\\/|\\)|\\:",replacement = "_")

mdFengQ_2015_SupplData1_3 <- mdFengQ_2015_SupplData1_2 %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,3:25]), as.integer)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,26:31]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,32:38]), as.integer)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,39:46]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,47]), as.numeric)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,48]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,49]), as.numeric)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_2[,40]), as.factor))

mdFengQ_2015_SupplData1_3 <- mdFengQ_2015_SupplData1_3 %>%
  mutate(Clinical_data_State = factor(Clinical_data_State, levels=c("controls", "advanced adenoma","carcinoma")))

#edit tables as some muberic are actually factors
mdFengQ_2015_SupplData1_4 <- mdFengQ_2015_SupplData1_3 %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,7:9]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,16:17]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,24:37]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,39:46]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,48]), as.factor)) %>%
  mutate(across(names(mdFengQ_2015_SupplData1_3[,50]), as.factor))

# I have now a well formatted table
mdFengQ_2015_SupplData1_4
write_csv(mdFengQ_2015_SupplData1_4,"curated-data/mdFengQ_2015_SupplData1_4.csv")

```